Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-28 paper/section-introduction.tex
Reordered intro a little; added paper structure section

Inspired by this connection, we study type systems that
capture rich invariants through types indexed by attributes with
algebraic structure.  For example, in computational geometry, points
in the plane can be indexed by attributes representing affine
transformations; in information-flow security, computations can be
indexed by principals; in differential privacy, types can be indexed
by `distance'. Types that are polymorphic over such indices induce
invariance properties and abstraction barriers beyond those introduced
by their unindexed versions, as we shall illustrate.  This generalises
previous work by the third author on types parameterized by
units of measure, whose invariance properties relate to changes of
units, or \emph{scaling}~\cite{kennedy97relational}.

Our motivations for studying algebraically indexed types are
threefold. First, we believe that, as with
units of measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

\paragraph{Invariance}
To illustrate type-induced invariance properties, consider
2-dimensional geometry. 
A function $\mathit{areaTri}$ that computes the area of a triangle
can be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But we can also assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
\]
This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
is translated by the same vector, then the result remains the same,
that is, it is \emph{invariant} under translation. Formally, for any 
vector $\vec t$,
\[
\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
\]

@@

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem, for a given \emph{model}
of index sorts and relational interpretation of primitive types.
We prove that the semantics soundly approximates contextual equivalence.
\item
For each of our main examples we deduce free theorems that are
consequences of our Abstraction Theorem, prove non-definability
results, and derive interesting type isomorphisms. \autoref{fig:summary}
illustrates the central position of our analogue of Reynolds' Abstraction
Theorem (\thmref{thm:abstraction}) in these results.
\item
We improve on the earlier semantics of
units of measure~\cite{kennedy97relational} in a number of ways.  By
extending the language of units with an `absolute value' operation, we
can give more precise types and obtain more general invariance
properties.  The relational interpretation for units is both simpler
and more flexible, and we derive slicker proofs of non-definability,
and new results.  Our notion of type isomorphism is stronger than
previously, being based on contextual equivalence.
\end{itemize}
We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

>>>>

Inspired by this connection, we study type systems that
capture rich invariants in types indexed by attributes with
algebraic structure.  For example, in computational geometry, points
in the plane can be indexed by attributes representing affine
transformations; in information-flow security, computations can be
indexed by principals; in differential privacy, types can be indexed
by `distance'. Types that are polymorphic over such indices induce
invariance properties and abstraction barriers beyond those introduced
by their unindexed versions, as we shall illustrate.  This generalises
previous work by the third author on types parameterized by
units of measure, whose invariance properties relate to changes of
units, or \emph{scaling}~\cite{kennedy97relational}.

\paragraph{Invariance}
To illustrate type-induced invariance properties, consider
two-dimensional geometry. 
In a conventional type system, a function $\mathit{areaTri}$ that computes the area of a triangle
might be assigned the type:
$\tyPrimNm{vec}\times\tyPrimNm{vec}\times\tyPrimNm{vec}\tyArr\tyReal$.
But in our proposed system we can assign it the following more expressive polymorphic
type:
\[
\mathrm{areaTri} : \forall t\mathord:\SynTransl{2}.
  \tyPrim{vec}{t} \times \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyReal \\
\]
This type expresses the fact that if each of the arguments to $\mathrm{areaTri}$
is translated by the same vector, then the result remains the same,
that is, it is \emph{invariant} under translation. Formally, for any 
vector $\vec t$,
\[
\mathrm{areaTri}\;(\vec{t} + \vec{v_1}, \vec{t} + \vec{v_2}, \vec{t} + \vec{v_3}) = 
\mathrm{areaTri}\;(\vec{v_1}, \vec{v_2}, \vec{v_3})
\]

@@

\paragraph{Motivations}
Our motivations for studying algebraically indexed types are
threefold. First, we believe that, as with
units of measure~\cite{fsharp}, practical programming language
extensions will follow. For example, in computational geometry and
graphics, attributes on points, vectors, and other geometric types
could be used to prevent the mixing of different coordinate systems,
or `frames'. Second, type-based static analyses can be based on
indexed types, for example, in effect systems~\cite{benton06reading},
and, more speculatively, in continuity
analysis~\cite{chaudhuri10continuity}.  Finally, we believe that
expressing algebraic invariants through types has the potential to
offer slick proof techniques for mechanized
mathematics. Harrison has applied the
invariance properties of geometric primitives to create elegant proofs
in geometry, based on `without loss of
generality' principles~\cite{harrison09without}. The invariance properties are expressed 
and propagated using ad-hoc tactics; our types offer a more principled
means of achieving the same end, and the `wlog' principle itself
is expressed through type isomorphisms.

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem, for a given \emph{model}
of index sorts and relational interpretation of primitive types.
We prove that the semantics soundly approximates contextual equivalence.
\item
For each of our main examples we deduce free theorems that are
consequences of our Abstraction Theorem, prove specific non-definability
results, and derive interesting type isomorphisms. 
For a large class of first-order types we give a general method for 
constructing suitable models to prove non-definability results. 
\autoref{fig:summary}
illustrates the central position of our analogue of Reynolds' Abstraction
Theorem (\thmref{thm:abstraction}) in these results.
\end{itemize}
We improve on the earlier semantics of
units of measure~\cite{kennedy97relational} in a number of ways.  By
extending the language of units with an `absolute value' operation, we
can give more precise types and obtain more general invariance
properties.  The relational interpretation for units is both simpler
and more flexible, and we derive slicker proofs of non-definability,
and new results.  Our notion of type isomorphism is stronger than
before, being based on contextual equivalence.

We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}
\subsection{Structure of paper}
Our paper is structured as follows. In \autoref{sec:motivating-examples}
we present an extended case study of two-dimensional geometry, with semi-formal
description of types and results.
In \autoref{sec:a-general-framework} we describe fully formally a general framework
for algebraically-indexed types, and prove the Abstraction Theorem and
soundness for semantic equivalence. \autoref{sec:instantiations} presents a number
of applications of the theory to 2-d geometry, including free theorems, type isomorphisms
and non-definability. \autoref{sec:general-nondef} develops a more general technique 
for proving non-definability results. \autoref{sec:information-flow} presents
the application of algebraically indexed types to information flow security, and
\autoref{sec:metric-types} applies it to types indexed by distances.
Finally, \autoref{sec:discussion} discusses related work and future plans.
